manuscript_v1

### 2022-02-11 manuscript.Rmd
manuscript v1.4.1 (moved tables & figures to end for Ibis guidelines)
manuscript v1.4.2 (cut text in discussion, summarising and gluing left)

This disproportionate attention is not unjustified: habitats chosen by animals can influence fitness outcomes via costs (e.g., predation risk) and benefits (e.g., food availability) of habitat use, and these have traditionally been strongly linked with reproductive success and offspring survival. In migrants, habitat requirements are complex and vary through the annual cycle, across their geographic range, along migratory routes, and at multiple spatial scales. Therefore, investigating their selection patterns may reveal a lot about the links between organisms and habitat and landscape features. Nevertheless, habitat dynamics during transitional extra-breeding periods can be of great importance as well, given their associated high stress and consequence as an important link in the chain. In the context of temperate birds, questions regarding habitat selection, niche breadth and community structure take on new dimensions in these periods. For one, the birds are less territorial and more mobile, allowing greater freedom in choosing foraging areas (Uesugi and Murakami, 2007). Additionally, motivations for habitat selection are straightforward and unambiguous; the main challenges are food scarcity coupled with increased food requirements, and the consequently higher predation risk---both for migrants that need to build up enough nutritional reserves and for residents that need to prepare for the oncoming winter. Parent-dependent fledgelings also transition to independent young birds that have to overcome challenges brought by the changing season and by their lack of experience in foraging, navigating and avoiding predators (Anders et al., 1997). Especially to them but also to adults, dense vegetation may provide cover and protection from adverse weather and predators. Dispersing juveniles may even collect social information regarding future site quality during the post-breeding period (Betts et al., 2008a). Despite slowly growing evidence that habitat selection in birds may differ seasonally due to the varying ecological requirements during the different seasons (Rivera et al., 1998; King et al., 2006; Akresh et al., 2009; Vitz and Rodewald, 2011; Dittmar et al., 2014), this period remains understudied.

@@

The final model was a GLMM with Poisson error distribution and log link function fitted with the BOBYQA optimiser (Table 2). It contained total bird abundance as response, and crossed random effects of sampling point and day. The predictors selected as important were observer (detections had 71:29 observer ratio), week, canopy heterogeneity, tree density, and dominant ground vegetation and its interaction with week.

@@

Here, it is worth noting that in the process of selection of candidate habitat variables, canopy heterogeneity correlated negatively with canopy cover at a significant level (r = -0.533, p = 0.0017). However, the relation was stronger at high values of canopy cover and weaker at low values, which is a direct physical result—with higher average cover, the potential variability in cover is lower. Additionally, tree density correlated significantly with canopy cover, albeit in the opposite direction (r = 0.526, p = 0.002). In spite of these two correlations, heterogeneity and tree density remained uncorrelated (r = -0.246, p = 0.1745). Further, tree density also correlated negatively with proportion of deciduous trees at a significant level (r = -0.545, p = 0.0013), but seemed to be the more important predictor in the models.

@@

Thirdly, to explore whether patterns differed between migrant and non-migrant species, the bird data was summarised by migration strategy and attempts were made to build GLMMs using the glmmTMB package. However, while a good fitting model was obtained for non-migrants (which had the highest number of overall detections), data for the other two categories seemed to be unsuitable and insufficient. The data showed significant one-inflation when tested (using functions in the DHARMa package) relative to both Poisson and negative binomial distributions, i.e., the number of ‘1’ count observations was too high for a standard Poisson or a standard negative binomial. This one-inflation in turn also caused underdispersion, which was removed neither on switching to distributions suited to underdispersed data (such as the Generalised Poisson and Conway-Maxwell Poisson), nor on introducing a zero-inflation term. The same problem occurred during attempts to build models at the species-level, i.e., looking at patterns of habitat selection for certain selected species (incidentally, the ones with most overall detections). The results of these questions were expected to be interesting, but as the data didn’t seem much suited to a binomial distribution either, this approach was abandoned for lack of better data. Finally, separate models were built to test whether bird abundances were linked to the phenology of the fruiting plants Rubus sp., Vaccinium sp., F. alnus and S. aucuparia but there was no detectable relationship.

The general increase in bird abundance seen with the progress of the season could be due to migrants arriving from the north to winter here, such as multiple tit species, but could also be due to better detection rates as the birds become bolder and more active, especially with the formation of large (and sometimes mixed) foraging flocks.

One of the biggest difficulties in studying bird communities in the non-breeding season is believed to be their low detectability. While we observed this to be true—birds were relatively very quiet, inconspicuous and inactive, particularly in PB—we realised that it was not as low as expected. The very low number of instances of birds feeding on fruits was unexpected. This could have been because arthropod abundance in the forest was still high enough to support the birds, or it could mean that birds in this region (mostly non-migrants or partial migrants) do not show the popular shift in foraging strategy that is seen elsewhere, particularly in North America (see Parrish, 1997, 2000), and rather only feed on fruits facultatively. However, since arthropod abundances were not measured in the study, it is hard to make definitive arguments.

The continuous measure of time (and seasonality) used in this study was considered to be better than comparing categorical factors like ‘post-breeding period’ and ‘pre-migratory period’, for a few reasons. Firstly, there is a fair degree of arbitrariness and ambiguity in published studies in this field regarding exact definitions of such seasons. Moreover, such categories might have very different meanings and implications depending on location. For instance, ‘pre-migratory periods’ in the northernmost parts of Scandinavia would be extremely different from those in, say, southern Czech Republic. This makes comparability difficult. Instead, measuring in numbers of weeks identified by a standard time unit (the Gregorian calendar) allows easy interpretability of results. And finally, such a measurement of time provides a much finer temporal resolution than would time categories, which could be important in spotting patterns developing at a relatively short time scale.

The overall abundance of forest birds after the breeding season seemed to be higher in forest patches where the canopy was highly heterogeneous and lower in forest patches with uniform canopy cover. There was also a negative trend with tree density despite the two habitat variables not being significantly correlated. Since a vast majority of bird detections were auditory and since such temperate forests with poor understorey layers do not limit auditory detections, the possibility of poorer detection rates with high tree density can be ruled out.

Heterogeneity has been explored before, but not in the same sense as in my study. For example, Freemark and Merriam (1986) observed positive trends of bird richness and abundance with overall habitat (rather than canopy) heterogeneity but also with forest size. They measured heterogeneity as a combination of different features such as species compositions of tree, shrub and ground strata and the forest structure. Their design consisted of multiple forest sites of varying areas and, naturally, heterogeneity was generally higher in larger forests. Hence, the results from their study do not relate well to those of mine, which concerned multiple points of the same area but of different qualities located within the same tract of forest.

One explanation for the trends then can be that both variables lead to higher light availability within the forest, which can be crucial for invertebrate-feeders particularly towards autumn when the conspicuous prey are all depleted. The trends did not interact with week, suggesting that these preferences are driven not by seasonally changing conditions (at least not at the scale I considered) but rather by some fundamental and more general driver. Canopy cover was not included in the models and hence its effect(s) cannot be compared. However, a potential reason that heterogeneity might be more important to birds than cover is that heterogeneous patches might provide better light conditions while still retaining at least some of the other desirable properties of the habitat. On the other hand, patches with very low average cover, despite having good light conditions, might not have sufficiently good vegetation characteristics.

In spite of the significant correlations between canopy heterogeneity and canopy cover, and between canopy cover and tree density, canopy heterogeneity and tree density remained uncorrelated (see Figure S6). This could be indicating two separable properties of canopy cover that often go unnoticed when comparing species responses such as abundance, richness or diversity with only canopy cover. Moreover, although the correlation between canopy cover and tree density is statistically significant, it seems weak at low densities of trees. This could mean that the trends with bird abundances are influenced not just by the availability of light. While heterogeneity may be allowing for better light conditions, low tree densities might be related to some other feature of the patch that is favourable to birds.

For example, Menon and Shahabuddin (2021) found that tree density across various sites in a low-latitude temperate forest, comprising both protected and managed patches, was highly negatively associated with mean tree girth, and observations suggested a negative relationship with canopy height as well. These could be due to suppressed regeneration and sapling growth caused by shade effects, and both variables were positively linked with bird abundance and richness (Menon and Shahabuddin, 2021). Although I did not measure canopy height and tree girth, the same trends can be expected here too. Wider trees can be expected to host a larger and more diverse group of arthropod prey species as a result of greater diversity of microhabitats, thereby favouring a higher abundance of birds.

Mixed forests with intermediate proportions of deciduous trees are generally believed to have higher biodiversity (e.g., Jansson and Andrén, 2003; Hagar, 2007; Ohno and Ishida, 1997; Felton et al., 2010). Alatalo (1981) observed that bird density increased linearly with proportion of deciduous trees in the summer, while the relative use of such habitats declined in winter. If not a linear relationship, bird abundances are expected to show a hump-shaped trend, with highest values found in mixed forests and lowest in both purely coniferous and purely broadleaved patches. However, tree density, with which the proportion of deciduous trees was correlated in my study, explained the variability in bird abundances better—neither a linear term nor a polynomial of the scaled proportion provided better fits.

Features of the understorey layer did not matter, even though the presence of shrubs in the understorey is generally expected to be good for small passerines. In my sampling points, the common species in the understorey were alder buckthorn and spruce, followed by pedunculate/sessile oak and birch. Other species occurred rarely and even when they did, were in very few numbers, sparse and isolated. This might explain the absence of a trend in bird abundance with understorey characteristics. These characteristics such as richness and diversity might not be captured with a wide enough range in my study area so as to matter for the birds. Hence, different patterns could be expected in other less-managed and naturally more diverse forests.

The dominant ground layer vegetation had a great effect on bird abundances, which was surprising because there is no reason to expect any special preference to one or the other species. However, these patterns could be merely reflecting some other feature that birds track. Rubus sp. and Vaccinium sp. ground layers were associated with similar and intermediate values of average bird abundance and increase in abundance, while moss and no-dominant (‘Bare’) ground layers were associated with similar values but had very low abundances in the beginning and very high values in the end. Graminoids like Carex spp. were associated with lowest average abundance as well as its increase.

In general, canopy heterogeneity, understorey density and understorey diversity were not particularly different between the different groups of points. There were weak patterns of the dominant layer with some habitat features: all groups except ‘Rubus’ had similar values of canopy cover while ‘Rubus’ had a lower mean and larger overall range of values; ‘Graminoid’ had slightly lower richness and diversity of canopy trees than others. The strongest patterns were seen in terms of coniferous/deciduous trees and tree density. The ‘Moss’ points (two in total) had the highest tree densities, a very low proportion of deciduous trees and a very high number of spruce trees, relative to all other points (Figure S7; see also Figure S2 A). Some of the ‘Bare’ points (i.e., with no dominant ground layer vegetation) also had high tree density and high numbers of spruce, but others had varied characteristics. ‘Rubus’ had the lowest tree densities, high proportions of deciduous trees and low spruce numbers. Others had mostly intermediate values. The relation with pine was similar to that with spruce, although ‘Moss’ and ‘Rubus’ did not have as extreme values. Finally, ‘Moss’ points had the lowest numbers of oaks and ‘Graminoid’ points had the highest, while others had intermediate numbers. A special case worth mentioning is that of Larix decidua. It occurred prominently in only two points, but according to my system was always considered as deciduous (and not as coniferous).

There seems to be a complex interplay of presence of conifers and density of trees. Several points other than ‘Bare’ and ‘Moss’, including one ‘Graminoid’ point, had relatively low proportions of deciduous trees but most of these were pine and not spruce. And, all other ‘Graminoid’ points had high numbers of oak. ‘Rubus’ and ‘Vaccinium’ points generally had mixed canopies as well as low tree densities. ‘Moss’ points had numerous spruces and very high tree densities. All this could be suggesting that in PB, birds preferred mixed forest patches that had some canopy heterogeneity and low densities of trees (which could in turn be related to tree girth, but also to proportion of deciduous trees). However, in PM, although such points still had relatively high abundances and hence can be considered as generally good for birds, an even greater increase in bird abundance was seen in points with highly dense canopy structure composed mainly of spruce. These spruce-dominated points were very poorly inhabited by birds in PB but had the highest abundances in PM.

Thus, while mixed species composition, canopy heterogeneity and low tree density are the best features generally, spruce-dominated dense patches represent an initially low-value niche that turns into a high-value niche. Pine, the other common conifer, did not have any major effect and was not as important as spruce was. ‘Graminoid’ points which had high oak numbers and generally low canopy heterogeneity were poorest for birds in both periods. This is a surprising result, as oaks are usually linked with higher biodiversity (especially Q. robur, which is highly insect-rich and has the highest visitation rates by birds), but can perhaps be better understood at a deeper level of response, i.e., at the level of guilds.

Alatalo (1981) observed that in general, mixed forests had the highest bird densities while pure pine forests and forests with low trees had the lowest bird densities. However, he also noticed that in winter, foliage-gleaners preferred pine over spruce, presumably due to the abundance of eggs of Lachnidae on pine needles. However, this is the opposite of what I observed, although it was not exactly in winter. Alatalo (1981) also suggests that species like Regulus regulus and Turdus spp. forage in pine forests in late summer, despite avoiding it during the nesting season. Berg (1997) noticed that the ground vegetation was related to tree species composition—‘grass’ vegetation had large volume of deciduous trees, which is in agreement with my observations of ‘Graminoid’ points, and Vaccinium spp. had more pine, which is different from what I observed. However, Berg also observed positive correlations of bird abundance with spruce and none with pine (similar to my result), but this was during the breeding season, when spruce is believed to be preferred over pine due to its special canopy structure that provides protected nest sites. Nevertheless, he suggests the possibility of differences in productivity and food availability being important criteria as well.

The reason the overall trend with tree density in the model is negative could be that the degree of avoidance of high-density spruce patches in PB might have been relatively high, and the subsequent preference in PM might not have been high enough to appear as an interaction of the variable. It became visible, however, when points were split into groups. Secondly, tree density provides a better fit than proportion of deciduous trees despite their significant correlation, probably because of the variability in values; the same value of tree density can arise in both conifer-dominated and deciduous broadleaf-dominated patches (see Figure S6). The two variables could have different implications to the birds and hence could be selected for differently.

The two guilds showing different patterns is indirect evidence of differences in their resource-use strategies. Further, while abundance of omnivores increased with time, that of invertebrate-feeders remained largely constant (as inferred from the lack of a notable effect of week in the model). This agrees with the fact that most of the northern migrants that arrive here are omnivores (mostly tits), and also suggests that detectability of the naturally quiet invertebrate-feeders was not a major issue in this study. If there actually was an increase in true abundance in PM, it would have to mean that detectability was higher in PB and lower in PM, which is highly unlikely.

Abundance of invertebrate-feeders decreased linearly with proportion of deciduous trees and was lower in graminoid-dominated points. Thus, they were clearly avoiding oak-dominated patches with low canopy heterogeneity, while their preferences were mostly similar among the other points regardless of whether they were spruce-dominated or mixed forest patches. However, tree density gave a poorer fit than proportion of deciduous trees, and this could be indicating the different implications of the two variables in spite of their correlation.

Invertebrate-feeders not preferring deciduous trees is surprising, given that oaks are generally considered to host good numbers of arthropod prey, but this could be because of a few different reasons. A majority of the herbivorous arthropods on oaks belong to Lepidoptera (45% of the total species richness of phytophages on Q. robur/petraea; Kennedy and Southwood, 1984), which attain a peak density in spring. The trees see relatively very low lepidopteran attacks after mid-June (Feeny, 1970). When considering all phytophagous insects, there is a second but much smaller peak in abundance in late summer and early autumn of suckers, leaf-miners and gall-formers (Southwood et al., 2004). Thus, the overwhelming preference for oaks by arthropods, and consequently by predatory birds, is likely to be prevalent mainly in spring and summer. The arthropod communities of oaks and their regulation by predators has rarely been explored in the post-breeding periods of late summer and early autumn.

Moreover, the specific invertebrate-feeding species present during this summer-autumn transition period such as treecreepers, robin and goldcrest/common firecrest are largely specialists and do not prey on external-feeding herbivores. Treecreepers are bark-feeders, robin is mostly a ground-forager and goldcrest is specialised for spruce (firecrest shows no preference for either conifers or broadleaf trees; Leisler and Thaler, 1982). In fact, even omnivores like the great tit have been shown to preferentially feed on conifers like pines (Eeva et al., 1997).

The model for omnivores had a negative binomial distribution, which is in concordance with the large foraging flocks seen among omnivore species such as tits. Abundance increased with time, likely due to the migrants, but average abundance was highest in forest edges. Abundance decreased with tree density like in the case of all birds, and there was no relationship with the dominant ground layer vegetation. Their preference of the structural feature (tree density) over the species’ biological characteristics like coniferous/deciduous suggests that being generalists, they are not particular about the food source. In the beginning, there was no major trend in abundance with canopy heterogeneity although the birds seemed to prefer less heterogeneous points. However, this was completely reversed by the end of the study, with the most heterogeneous points having the highest abundance—twice that of the least heterogeneous points which did not see any change in bird abundance at all over the course of the season.

Although invertebrate-feeders did not show preference to or against heterogeneous patches, omnivores seemed to have avoided them in the early weeks, the reason for which is not obvious. The subsequent arrival of the migrants caused an occupation of the previously ‘empty’ niche and this was most likely because heterogeneity enables the formation of large foraging flocks, with different individuals in the party focussed on different aspects of the habitat. This, in combination with higher light availability, would result in higher foraging efficiency, which is crucial in the pre-migratory period. The fact that the dominant ground vegetation was not important to omnivore abundance especially with the arrival of the migrants further solidifies that structural features were the only things they tracked, and that they were otherwise rather evenly distributed over the study area.

Wolfe et al. (2014) suggested that because forest-dwelling arthropod species are often cryptic (Bond, 2007), it may be inherently difficult and inefficient for an insectivore to directly assess the availability of arthropods, particularly in resource-scarce periods. They suggested this to be the reason why frugivores and insectivores tracked different cues. However, in my study, the species belonging to the ‘invertebrate-feeder’ group are naturally specialised on cryptic prey species. This could be why they tracked the qualitative features of the food resources (even if not directly but by the coniferous/deciduous distinction) over structural characteristics. On the other hand, the generalist omnivores would have had difficulty foraging for these cryptic species, and hence attained higher foraging efficiency by tracking structural characteristics, especially with the formation of foraging flocks.

## Post-breeding and pre-migratory behaviour of bird species

The dissimilarity in niches occupied by the two guilds is evident in the ordinations too. Even among omnivores, the divergent Dendrocopos spp. might be showing a preference for deciduous trees, in that it occurred in points with intermediate proportions of deciduous trees and intermediate tree densities. Thus, for them, the biological trait of coniferous/deciduous was well-linked with the associated structural characteristic of density. While hole-nesting birds, such as woodpeckers, are known to prefer deciduous trees for breeding and for better predator protection, in my study this seemed to be also the case in both PB and PM. Despite being omnivorous, they occupied a different niche than the small omnivorous passerines like tits and nuthatch, which makes sense due to their vastly different foraging strategies. The nuthatch preferences were slightly different from those of tits in PB while they converged in PM, facilitating the mixed-species foraging flocks.

On the other hand, with the exception of treecreepers, invertebrate-feeders showed a divergence in strategies in the PM. This was possibly due to the shortage of prey in this period, which would have increased competitive pressure, forcing them to diverge. This is most notable in the case of Regulus spp. and Phylloscopus spp. which are both foliage-gleaners, although the feature they tracked most was still tree density. Robin and treecreepers appeared to expand slightly into more deciduous patches in PM relative to PB. Interestingly, although the statistical models were unable to pick up any major trends with understorey features, the ordinations seem to suggest that there were indeed some, particularly for invertebrate-feeders and in PB. This further solidifies the possibility that other more ‘natural’ and less-manipulated forests with a much wider range of understorey features than seen in Branišovský les could show different patterns.

Overall, the bird communities defined by these eight species across the three habitat classes of my study points were more similar in PM, suggesting a relatively generalist strategy. In PB, the communities of the three habitat classes were more separated than in PM, when they overlapped greatly.

## Bird predation on plasticine caterpillars

In general, not many studies have looked at changes in predation rates with season (but see e.g., Molleman et al., 2016 in the tropics). There are two main assumptions that can be made in interpreting the results from my study. First, predation rates on the dummy caterpillars can be assumed to be associated with the true predation rates (i.e., predation on actual arthropod prey) prevalent at that particular time. Or second, the plasticine predation rates can be assumed to not reflect accurately true predation rates. The method of studying predation rates using artificial caterpillars fails to accurately match the natural predation rates and are usually conservative (Howe et al., 2009), although not many have studied this question in the summer-autumn transition in temperate regions. Nevertheless, the use of model prey permits a standardised, rapid assessment of relative predation rates across the seasonal gradient and between different points.

Thus, the plasticine predation only approximates true predation. However, there are two further assumptions that can be made. This link (under- or overestimated approximation) can be assumed to be with either the general arthropod predation rates, or specifically lepidopteran predation rates. The latter appears to be most reasonable, as the artificial caterpillars used in the study were made to resemble generic lepidopteran larvae of oaks in Central Europe. Moreover, there was no evident shift towards frugivory (as inferred from the very low fruit foraging observations relative to overall foraging observations [2/494 = 0.4%]) despite the decline in predation. So, the former assumption can be discarded.

It is then safe to assume that birds are able to distinguish, to some extent, between artificial and natural caterpillars. Additionally, simply by virtue of probabilities, predation on artificial prey correlates negatively with availability of natural prey (data of Volf et al., 2021). Thus, an increase can be expected in predation on artificial prey with a significant decline in natural prey availability. In my study, bird predation on artificial caterpillars declined considerably over the weeks and was affected to a much lesser extent by habitat class. Other habitat variables did not have a detectable impact in my study area. This decline could be either because natural prey availability increased, or because birds lost interest in the artificial prey over the course of the season in spite of an unchanged natural prey availability.

>>>>

This disproportionate attention is not unjustified: habitats chosen by animals can influence fitness outcomes via costs (e.g., predation risk) and benefits (e.g., food availability) of habitat use, and these have traditionally been strongly linked with reproductive success and offspring survival. In migrants, habitat requirements are complex and vary through the annual cycle, across their geographic range, along migratory routes, and at multiple spatial scales. Therefore, investigating their selection patterns may reveal a lot about the links between organisms and habitat and landscape features. Nevertheless, habitat dynamics during transitional extra-breeding periods can be of great importance as well, given their associated high stress and consequence as an important link in the chain. In the context of temperate birds, questions regarding habitat selection, niche breadth and community structure take on new dimensions in these periods. For one, the birds are less territorial and more mobile, allowing greater freedom in choosing foraging areas (Uesugi and Murakami, 2007). Additionally, motivations for habitat selection are straightforward and unambiguous; the main challenges are food scarcity coupled with increased food requirements, and the consequently higher predation risk---both for migrants that need to build up enough nutritional reserves and for residents that need to prepare for the oncoming winter. Parent-dependent fledgelings also transition to independent young birds that have to overcome challenges brought by the changing season and by their lack of experience in foraging, navigating and avoiding predators (Anders et al., 1997). Especially to them but also to adults, dense vegetation may provide cover and protection from adverse weather and predators. Dispersing juveniles may even collect social information regarding future site quality during the post-breeding period (Betts et al., 2008a). Things are not made easier by the arrival of migrants from the north, resulting in a higher overall resource consumption. Despite slowly growing evidence that habitat selection in birds may differ seasonally due to the varying ecological requirements during the different seasons (Rivera et al., 1998; King et al., 2006; Akresh et al., 2009; Vitz and Rodewald, 2011; Dittmar et al., 2014), the post-breeding period remains understudied, at least in part due to the much lower vocal activity of birds in this period.

@@

The final model was a GLMM with Poisson error distribution and log link function fitted with the BOBYQA optimiser (Table 2). It contained total bird abundance as response, and crossed random effects of sampling point and day. The predictors selected as important were observer (detections had 71:29 observer ratio), week, canopy heterogeneity, tree density, and dominant ground vegetation and its interaction with week. In spite of the significant correlations between canopy heterogeneity and canopy cover (r = -0.533, p = 0.0017), and between canopy cover and tree density (r = 0.526, p = 0.002), canopy heterogeneity and tree density remained uncorrelated (r = -0.246, p = 0.1745). Further, tree density also correlated negatively with proportion of deciduous trees at a significant level (r = -0.545, p = 0.0013), but seemed to be the more important predictor in the models.

@@

While our observations corroborated the general understanding that birds are more quiet, inconspicuous and inactive during the post-breeding period, detectability was not as low as expected. The general increase in bird 'abundance' seen with the progress of the season could have been due to migrants arriving from the north to winter here, such as multiple tit species, but also due to better detection rates as the birds became bolder and more active, especially with the formation of large (and sometimes mixed) foraging flocks.

The very low number of instances of birds feeding on fruits was unexpected but solidifies the distinction between Central European and North American forests. Either arthropod abundance in the forest was still high enough to support the birds, or the birds in this region---mostly non-migrants or partial migrants---simply do not shift their foraging strategy, unlike birds elsewhere (see Parrish, 1997, 2000), and rather only feed on fruits facultatively.

We used a continuous measure of time (and seasonality) rather than categorical factors like ‘post-breeding period’ and ‘pre-migratory period’, for better comparability. Firstly, there is high ambiguity and arbitrariness in the literature regarding exact definitions of such seasons. Moreover, such categories would have very different meanings and implications depending on location. For instance, ‘pre-migratory periods’ in the northernmost parts of Scandinavia would be extremely different from those in, say, southern Czech Republic. Instead, measuring in numbers of weeks identified by a standard time unit (the Gregorian calendar) allows easy interpretation of results. Finally, such a measurement of time provides a much finer temporal resolution, which could be important in spotting patterns developing at a relatively short time scale. That said, other factors affecting comparability such as choice of bird survey method and habitat variables persist in our study.

The overall abundance of forest birds after the breeding season seemed to be higher in forest patches where the canopy was highly heterogeneous and lower in forest patches with uniform canopy cover. There was also a negative trend with tree density despite the two habitat variables not being significantly correlated. Since a vast majority of bird detections were auditory and since such temperate forests with poor understorey layers do not limit auditory detections, the possibility of poorer detection rates with higher tree density can be ruled out, although it is unknown whether tree density also affects overall bird behaviour. Although the presence of shrubs in the understorey is generally expected to be good for small passerines, in our study, the common species in the understorey were *F. alnus* and *P. abies*, followed by *Q. robur/petraea* and *B. pendula*. Other species occurred rarely and even when they did, were sparse, isolated and in very few numbers. This likely explains the absence of a trend in bird abundance with understorey characteristics such as richness and diversity, which might not be captured with a wide enough range in my study area so as to matter for the birds. Hence, different patterns could be expected in other less-managed and naturally more diverse forests.

Heterogeneity has been explored before in habitat selection studies, but the focus has mostly been on the habitat and not the canopy. For example, Freemark and Merriam (1986) observed positive trends of bird richness and abundance with overall habitat heterogeneity (combination of species compositions and forest structure) and forest size, which are expected to correlate. Canopy heterogeneity however, along with canopy cover, affect light availability within the forest, which can be crucial for invertebrate-feeders, particularly in autumn when the conspicuous prey are depleted. In our study, across multiple points of the same area within the same forest tract, the trends between bird abundance and canopy heterogeneity did not change with time, suggesting that these preferences are driven not by seasonally changing conditions but rather by some fundamental and more general driver. *Canopy cover was not included as a candidate predictor and hence its effect cannot be compared*, but heterogeneity might potentially be more important to birds as heterogeneous patches provide better light conditions while still retaining some other desirable properties of the habitat. Patches with low average cover, despite having good light conditions, might not have sufficiently good vegetation characteristics. This could be indicating two separable properties of the canopy that often go unnoticed.

On the other hand, the correlation between canopy cover and tree density, despite being statistically significant, is weak at low densities, suggesting that the preference of birds for low density patches might be driven by some factor other than light availability. For instance, Menon and Shahabuddin (2021) found that tree density across various sites in a low-latitude temperate forest, comprising both protected and managed patches, was highly negatively associated with mean tree girth, and observations suggested a negative relationship with canopy height as well. While these could be due to suppressed regeneration and sapling growth, and therefore indirectly due to light availability (shade effects), both variables were positively linked with bird abundance and richness (Menon and Shahabuddin, 2021). Wider and taller trees are expected to host a larger and more diverse group of arthropod prey species as a result of greater diversity of microhabitats.

Birds generally show a hump-shaped trend with proportion of deciduous trees, with highest numbers found in mixed forests and lowest in both purely coniferous and purely broadleaved patches (e.g., Alatalo, 1981; Jansson and Andrén, 2003; Hagar, 2007; Ohno and Ishida, 1997; Felton et al., 2010). However, in alignment with Alatalo's findings (1981) that relative use of such deciduous patches were strong in summer but declined in winter, birds in our study in the summer-winter transition showed neither a linear nor a polynomial relationship with proportion of deciduous trees.

*probably continue with omn ins split here*

Surprisingly, the dominant ground layer vegetation had a great effect on bird abundances. Canopy heterogeneity, understorey density and understorey diversity were not particularly different between the different groups of points, and there were only weak patterns with canopy cover, richness and diversity. However, as observed in a central Swedish forest by Berg (1997), the ground vegetation seems to be related  to tree species composition. Specifically, there is a complex interplay of conifer presence and tree density. The ‘Moss’ points (two in total), relative to all other points, had the highest tree densities, very low deciduous proportions (lowest oak numbers) and very high spruce numbers (Figure S7; see also Figure S2 A). Some of the ‘Bare’ points (no dominant ground layer vegetation) also had high tree densities and high numbers of spruce, although others had varied characteristics. Bird abundance in both 'Moss' and 'Bare' points was very low in the beginning and very high in the end. Several points other than 'Moss' and 'Bare', including one ‘Graminoid’ point, had relatively low proportions of deciduous trees but these consisted mostly of pine and not spruce. While the relation of different ground layers with pine was similar to that with spruce, it was not as strong. ‘Rubus’ and ‘Vaccinium’ points generally had mixed canopies with high deciduous proportions and low spruce numbers as well as low tree densities ('Rubus' had the lowest). Such points had intermediate values of both average bird abundance and increase in abundance. Finally, while other points had intermediate oak numbers, 'Graminoid' points had the highest (like the 'grass' vegetation in Berg, 1997) and these were also associated with generally low canopy heterogeneity. These points had the lowest average bird abundance as well as its increase.

Thus, birds early in the season preferred mixed forest patches that had some canopy heterogeneity and low tree densities (which could in turn be related to either tree girth or deciduous proportions), and avoided spruce-dominated points. However, closer to autumn, these points with highly dense canopy structure were increasingly utilised and preferred, and thus had highest bird abundances in the end. All this might be suggesting that although mixed species composition, canopy heterogeneity and low tree density are generally the best features and hence can be considered as generally good for birds (such points had relatively high abundances in the end too), spruce-dominated dense patches represent an initially low-value niche that turns into a high-value niche. While many studies (e.g., Alatalo, 1981) directly compare the impacts of tree density and deciduous proportion as habitat variables, they could have different implications to birds and hence could be selected for differently. In the final model, tree density provides a better fit than proportion of deciduous trees despite their significant correlation, likely due to the variability in values; the same value of tree density can arise in both conifer-dominated and deciduous broadleaf-dominated patches (see Figure S6). Moreover, the overall trend with tree density in the model is negative possibly because the degree of initial avoidance of high-density spruce patches was relatively high while the subsequent preference was not high enough to appear statistically as an interaction of the variable. It became visible, however, when the sampling points were split into groups.

Interestingly, the same effect was absent with pine, the other common conifer, which could mean it is less important than spruce in our study area. Berg (1997) similarly observed positive correlations of bird abundance with spruce and none with pine, but this was during the breeding season, when spruce is believed to be preferred over pine due to its special canopy structure that provides protected nest sites, and possibly also due to differences in productivity and food availability. On the other hand, Alatalo (1981) noticed that foliage-gleaners preferred pine over spruce, presumably due to the abundance of eggs of Lachnidae on pine needles, but this was in winter. However, he also suggests that species like firecrest and *Turdus* thrushes forage in pine forests in late summer, despite avoiding it during the nesting season.

Oaks are usually linked with higher biodiversity (especially *Q. robur*, which is highly insect-rich and has the highest visitation rates by birds), but oak-rich points in our study being poorest for birds in both periods can perhaps be better understood at a deeper level of response, i.e., at the level of guilds.

The two guilds showing different patterns is indirect evidence of differences in their resource-use strategies. Further, while abundance of omnivores increased with time, that of invertebrate-feeders remained largely constant. This agrees with the fact that most of the northern migrants that arrive here are omnivores (mostly tits), also suggesting that detectability of the naturally quiet invertebrate-feeders was not a major issue in this study.

Abundance of invertebrate-feeders decreased linearly with proportion of deciduous trees and was lower in graminoid-dominated points. Thus, surprisingly, they were clearly avoiding oak-dominated patches with low canopy heterogeneity, while their preferences were mostly similar among the other points regardless of whether they were spruce-dominated or mixed forest patches. It is to be noted that a majority of the herbivorous arthropods on oaks belong to Lepidoptera (45% of the total species richness of phytophages on *Q. robur/petraea*; Kennedy and Southwood, 1984), which attain a peak density in spring. The trees see relatively very low lepidopteran attacks after mid-June (Feeny, 1970). When considering all phytophagous insects, there is a second but much smaller peak in abundance in late summer and early autumn of suckers, leaf-miners and gall-formers (Southwood et al., 2004). Moorman et al. (2007) have shown that lepidopterans are less preferred than coleopterans and hymenopterans. Thus, the overwhelming preference for oaks by arthropods, and consequently by predatory birds, is likely to be prevalent mainly in spring and summer. The arthropod communities of oaks and their regulation by predators has rarely been explored in the post-breeding periods of late summer and early autumn. Moreover, the specific invertebrate-feeding species present during this summer-autumn transition period such as treecreepers, robin and goldcrest/common firecrest are largely specialists and do not prey on external-feeding herbivores. Treecreepers are bark-feeders, robin is mostly a ground-forager and goldcrest is specialised for spruce (firecrest shows no preference for either conifers or broadleaf trees; Leisler and Thaler, 1982). In fact, even omnivores like the great tit have been shown to preferentially feed on conifers like pines (Eeva et al., 1997).

The model for omnivores had a negative binomial distribution, corroborating the presence of large foraging flocks seen among omnivore species such as tits. Abundance increased with time, likely due to the migrants, but average abundance was highest in forest edges. Abundance decreased with tree density and there was no relationship with the dominant ground layer vegetation. Their structural preference (tree density) rather than biological or taxonomical preference (like coniferous/deciduous) suggests that being generalists, they are not particular about the nature of their food source. In the beginning, like invertebrate-feeders, there was no major trend in omnivore abundance with canopy heterogeneity although the birds seemed to prefer less heterogeneous points. However, this was completely reversed by the end of the study, with the most heterogeneous points having the highest abundance---twice that of the least heterogeneous points which did not see any change in bird abundance at all over the course of the season. The arrival of migrants caused an occupation of the previously ‘empty’ niche and this was most likely because heterogeneity enables the formation of large foraging flocks, with different individuals in the party focussed on different aspects of the habitat. This, in combination with higher light availability, would result in higher foraging efficiency, which is crucial in the pre-migratory period. Heterogeneity could also explain higher abundance in forest edges. The fact that the dominant ground vegetation was not important, especially with the arrival of the migrants, further solidifies that omnivores tracked only structural features, and that they were otherwise rather evenly distributed over the study area.

Wolfe et al. (2014) suggested that because forest-dwelling arthropod species are often cryptic (Bond, 2007), it may be inherently difficult and inefficient for an insectivore to directly assess the availability of arthropods, particularly in resource-scarce periods. They suggested this to be the reason why insectivores, unlike frugivores, tracked structural cues. However, our ‘invertebrate-feeding’ species are naturally specialised on cryptic prey species. This could be why they tracked the qualitative features of the food resources (indirectly by the coniferous/deciduous distinction) over structural characteristics. On the other hand, the generalist omnivores would have had difficulty foraging for these cryptic species, and hence attained higher foraging efficiency by tracking structural characteristics, especially with the formation of foraging flocks.

@@
